---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)

options(width = 80)
```

# mermaidrcovariates

<!-- badges: start -->
<!-- badges: end -->

The goal of `mermaidrcovariates` is to easily access covariates data and add it to MERMAID data.

For more information and detailed instructions on usage, please visit the [package website](https://data-mermaid.github.io/mermaidr-covariates/).

For more details on using `mermaidr` in general, please see the [`mermaidr` package website](https://data-mermaid.github.io/mermaidr/).

## Installation

You can install `mermaidrcovariates` from GitHub with:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("data-mermaid/mermaidr-covariates")
```

## Usage

Through `mermaidrcovariates`, you can easily add covariates to the aggregated data from your coral reef surveys, which is already entered in MERMAID. To do this, first load the `mermaidrcovariates` and `mermaid` packages, and access the sample events for a given project. For this example, we will work with hard coral cover data, available from the Benthic PIT method. 

```{r get-sample-events}
library(mermaidrcovariates)
library(mermaidr)

se <- mermaid_search_my_projects("Great Sea Reef 2019") %>%
  mermaid_get_project_data("benthicpit", data = "sampleevents", limit = 10)
```

To get covariates, we need each sample event's date as well as its latitude and longitude. We will keep these columns along with the site name and average hard coral cover, using the `tidyverse` package for data manipulation.

```{r se-cols}
library(tidyverse)

se <- se %>%
  select(site, sample_date, latitude, longitude, hard_coral_cover = percent_cover_benthic_category_avg_hard_coral)
```

You can use the [Prescient browser](https://mermaid.prescient.earth/) to find which covariates are available. To see this programatically, we use the `list_covariates()` function:

```{r list-covariates}
list_covariates()
```

For this example, we will look at maximum "Daily Sea Surface Temperature" (SST). We can access this data by using the function `get_summary_zonal_statistics()`, which takes the site latitude and longitude, as well as the survey date, to find the data at that site for `n_days` days prior, within the given `radius`, and *spatially* aggregates it according to the specified `spatial_stat`. Then, the function *temporally* aggregates it according to the specified `temporal_stat`.

For example, to get the **maximum** SST over the 20 days prior to (and including) the sample event, using the *mean* SST within 100m of the sites:

```{r get-summary-zonal-stats}
max_sst <- se %>%
  get_summary_zonal_statistics("Daily Sea Surface Temperature", n_days = 10, radius = 100, spatial_stats = "mean", temporal_stats = "max")
```

The covariates are returned in a format that need to be expanded. Once they are, you can see they contain start and end date of the data used for the covariates, the band, and the summarised value.

```{r output-width, echo = FALSE}
options(dplyr.width = Inf)
```

```{r summary-zonal-stats-unnest}
max_sst %>%
  unnest(covariates)
```

If we don't want the data aggregated over time -- for example, if we just want the value of a covariate the day of the sample event, or if we want to have the individual dates attached to the covariate data -- we use the non-summary version of the function, `get_zonal_statistics()`. In this case, we omit `temporal_stats` and just get the mean SST within 100m of the sites, for all 20 days:

```{r get-zonal-stats}
sst_by_day <- se %>%
  get_zonal_statistics("Daily Sea Surface Temperature", n_days = 10, radius = 100, spatial_stats = "mean")

sst_by_day %>%
    unnest(covariates)
```

Here, instead of having the data aggregated over time, there is one row for each date, along with the value on that date.
