---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)

options(width = 80)
```

# mermaidrcovariates

<!-- badges: start -->
<!-- badges: end -->

The goal of `mermaidrcovariates` is to easily access covariates data and add it to MERMAID data.

For more information and detailed instructions on usage, please visit the [package website](https://data-mermaid.github.io/mermaidr-covariates/).

For more details on using `mermaidr` in general, please see the [`mermaidr` package website](https://data-mermaid.github.io/mermaidr/).

## Installation

You can install `mermaidrcovariates` from GitHub with:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("data-mermaid/mermaidr-covariates")
```

## Usage

### `mermaidr` data

Through `mermaidrcovariates`, you can easily add covariates to the aggregated data from your coral reef surveys, which is already entered in MERMAID. To do this, first load the `mermaidrcovariates` and `mermaid` packages, and access the sample events for a given project. For this example, we will work with sample events available from the Benthic PIT method. 

```{r get-sample-events}
library(mermaidrcovariates)
library(mermaidr)

se <- mermaid_search_my_projects("Great Sea Reef 2019") %>%
  mermaid_get_project_data("benthicpit", data = "sampleevents", limit = 10)
```

To get covariates, we need each sample event's date as well as its latitude and longitude. We will keep these columns along with the site name, using the `tidyverse` package for data manipulation.

```{r se-cols}
library(tidyverse)

se <- se %>%
  select(site, sample_date, latitude, longitude)
```

### Covariates

You can use the [Prescient Browser](https://mermaid.prescient.earth/) to find which covariates are available. To see this programatically, we use the `list_covariates()` function:

```{r list-covariates}
covariates <- list_covariates()
covariates
```

Covariates may contain raster data, vector data, or both. There are different functions for accessing raster or vector data.

The table below summarises which function to use for the different covariates available on Prescient:

```{r covariates-fxn-summary, echo = FALSE}
covariates <- list_covariates()

covariates_type <- covariates %>%
  filter(!str_starts(title, "GPW")) %>%
  select(Covariate = title, id) %>%
  distinct() %>%
  rowwise() %>%
  mutate(Type = mermaidrcovariates:::get_collection_type(id)) %>%
  ungroup()

covariates_type %>%
  mutate(
    `get_zonal_statistics()` = ifelse(str_detect(Type, "raster"), "✅", ""),
    `attach_covariate_data()` = ifelse(str_detect(Type, "vector"), "✅", "")
  ) %>%
  select(-id) %>%
  knitr::kable()
```

You can also use the function `covariate_helper()` which will give you information on the data type and which function to use, as well as any other specifications that might be needed.

### Raster data

For example, we will look at "Daily Global 5km Satellite Sea Surface Temperature (CoralTemp)" (SST). Running the helper function tells us to use `get_zonal_statistics()`, and that we do not need to specify the dataset or band, since there are only one of each.

```{r daily-sst-helper, message = TRUE}
covariate_helper("Daily Global 5km Satellite Sea Surface Temperature (CoralTemp)")
```

If we want the maximum SST at each site over the past 30 days, we first get the *daily* SST using `get_zonal_statistics()`, specifying `n_days = 30`. We will request all data within a `radius` of 1000 metres from the site, and that the function *spatially* aggregates that data for each of the 30 days.

```{r daily-sst}
daily_sst <- get_zonal_statistics(se,
  covariate = "Daily Global 5km Satellite Sea Surface Temperature (CoralTemp)",
  spatial_stats = "mean", radius = 1000, n_days = 30
)

daily_sst
```

You can see that the results for each sample event contains 30 rows -- one value for each of the 30 days we requested.

To expand the zonal statistics, we use `unnest()`, and are able to see that we have the mean value for each date:

```{r output-width, echo = FALSE}
options(dplyr.width = Inf)
```

```{r daily-zonal-stats-unnest}
daily_sst %>%
  select(site, sample_date, zonal_statistics) %>%
  unnest(zonal_statistics)
```

Since we would like the *maximum* SST for each site over those 30 days, we then use `aggregate_zonal_statistics()`, specifying that the *temporal* summary statistic should be max:

```{r summary-zonal-stats}
max_sst <- daily_sst %>%
  summarise_zonal_statistics("max")

max_sst
```

We retain the original `zonal_statistics` column, which contains the daily values, but now also have a `summary_zonal_statistics` column, which contains only *one* row for each sample event. 

We expand it and see a summary of the start and end date of the data used for each sample event, as well as the spatial and temporal statistics used to summarise, and the actual value: 

```{r summary-zonal-stats-expanded}
max_sst %>%
  select(site, sample_date, summary_zonal_statistics) %>%
  unnest(summary_zonal_statistics)
```

### Vector data

If we want to attach vector data, such as Country Boundaries, we use `attach_vector_data()`. Again, the `covariate_helper()` function is useful for determining what to do here:

```{r covariate-helper-countries, message = TRUE}
covariate_helper("Country Boundaries")
```

This tells us which function to use, and that we do not need to specify the dataset for the covariate. It also tells us that, by default, all of the columns in the vector data will be returned, but that we can limit to specific columns.

We attach the covariate data:

```{r country-all-cols}
se_country_all_cols <- se %>%
  attach_covariate_data("Country Boundaries")

se_country_all_cols
```

In this case, it is *not* a format that needs to be expanded -- the columns are immediately available.

If we only want one column (e.g. the country name), then we can specify that in the function:

```{r country-only}
se_country <- se %>%
  attach_covariate_data("Country Boundaries", columns = "TERRITORY1")

se_country
```
